Creating More Meaningful Explanations for Audio Deepfake Detection

### 2025-02-20 draft.tex
typo
sort comparison table
general formulation changes

					Intensity: Also inspired by previoous research \cite{li_comparative_2022},

@@

			We trained fhe black-box model described in the previous section on the first portion

@@

			1,000,000 lines per batch. For each batch, either one or two epochs were performed.

@@

			the explanations cannot be generated directly from the input itself. However, because

@@

			the input of the terminus model would be. Intermediate data was generated by first

@@

			terminus, and for each sub-model, running a prediction on the given input features.
			The results of the predictions of the sub-models were stored and used for further
			analysis. This was done not only for the sample explanation data but also for the
			training data, to use as an input for the LIME explainer.

@@

			The results of the LIME explainer are then summarized together on a per-feature basis
			and normalized to produce a decimal number between -1 and 1 for every feature. In this
			case, a negative value implies that the given feature pushed the result of the model
			in the negative direction (i.e. not fake), while a positive value indicates the

@@

			previously described method together in a meaningful way, in order to make a
			conclusion about the general usefulness of the individual features and the influence
			and potential use of the perceptible features. However, due to a lack of existing
			standardized metrics for the global evaluation of local metrics, we propose two
			metrics, with which we will make conclusions about the usefulness of this method.

@@

	testing the surrogate model, as well as generating many explanations for the purpose of
	aggregate evaluation. This section will first discuss the results and performance of the
	surrogate model, followed by an evaluation of the aggregated explanations.

@@

		The dataset used for these experiments is the "In-the-Wild" dataset, one which focuses on

@@

		aforementioned focus on generalization, its relative recency, compared to some others and

@@

		A summary of some experiments based on the method described above is provided in table
		\ref{table:eval-results}. The table covers the type of terminus model used, where a
		"simple" terminus model does not include hidden layers, the set of features used, where
		"standard" is the feature set as described above including only one pitch-fluctuation
		feature, with expanded pitch fluctuations being the same feature set with the addition of
		more pitch fluctuation features with different comparison distances, the training batch
		size and the number of epochs per batch if there were more than one, followed by several
		metrics for comparison. Figures \ref{fig:roc_cterm}, \ref{fig:roc_mpf} and
		\ref{fig:roc_mpf_cterm} depict the ROC curve of three of the models, which were selected
		for use in explanation generation in the following section.

@@

			In order to contextualize our results, we will compare them to other results produced
			using tests on the same dataset. Table \ref{table:other_results} presents some
			experimental results that other methods have achieved on this dataset, sorted by
			source and model architecture. The results are given in EER, and the best result is in
			bold. In comparison to the other methods, our method performs very well, only beaten
			out by one other \cite{ranjan_statnet_2022}, which uses an expanded RawNet2-based
			network, taking the raw waveform as an input. They trained their method on 70\% of the
			dataset, used 10\% for validation and the remaining 20\% for testing. This indicates a
			potential for raw-waveform based learned feature in the accurate detection of audio
			deepfakes, but does not offer a clear path for explainability, as the model does not
			make any intrinsic distinction of audio characteristics that humans are able to perceive.
			Our model remains, however, competitive, indicating that it can provide a good base
			for trustworthy explanations.

@@

		on the first 500 samples of the testing data. The models selected were the best-performing
		model, without expanded pitch-fluctuation features and with three hidden layers, the
		model with expanded pitch-fluctuation features and a simple terminus, as well as the model
		with expanded pitch-fluctuation features and a terminus with three hidden layers. These
		models were selected because we wanted to firstly evaluate the best performing model, as
		well as see how the changed pitch-fluctuation feature affected the evaluation of the
		explanations. We then summarized the explanations generated using both of the previously

@@

		presented in tables \ref{table:exp-results-cterm}, \ref{table:exp-results-more-pitch-flucs}
		and \ref{table:exp-results-mpf-cterm}. The same statistics are also represented graphically
		in the figures \ref{fig:exp_cterm}, \ref{fig:exp_mpf} and \ref{fig:exp_mpf_cterm}
		respectively.

@@

		features had either little, no or negative influence on the classification.

@@

		more harm than good in the tested classifications.

@@

		examples, the trust value of the intensities feature was negative here.

@@

		for something more presentable to an end user. As can be seen in the previous section in

@@

		net-negative impact on the correctness of the classifications in the experiments, leaving
		the question open, if future experiments may perform better without these features. In the
		case of the best performing model, four perceptible features, f0 lengths, onset strengths,
		intensities and pitch fluctuations had an overall average influence on the classification
		of 0, measured on both metrics, indicating that the model did not learn a correlation
		between these features and the authenticity of audio samples in the dataset. These features
		did have an impact on the performence of the model with expanded pitch fluctuations, but
		the difference was still minimal. In the case of perceptible features with a positive
		trust score, the potential remains for their potential use in understandable explanations
		as, even though the classification needle of the model was not moved largely by these
		features, they still demonstrate a certain reliability. In spite of this, we believe that
		it is still worth investigating if other kinds of perceptible features will have more of
		an influence on the overall classification or if there are other imperceptible features
		that could be used in the place of MFCCs that may provide a better balance between
		classification correctness and a proportional influence on the end result, so that the
		usefulness of the explanations can be even stronger than the method using the above
		feature combinations, and we believe that this kind of model architecture paired with LIME
		and evaluated with the presented metrics has the potential to be a backbone for future
		research.

@@

		usability of such explanations open. Finally, we have not addressed in our experiments how

@@

	consistent and correct classifications. The explanations, however, were not able to produce
	the desired result, placing most of the weight on MFCCs, with most perceptible features either
	having little, no or negative impact on the end classification with the exception of jitter
	and shimmer, which offer potential to be used in explanations.

>>>>

					Intensity: Also inspired by previous research \cite{li_comparative_2022},

@@

			We trained the black-box model described in the previous section on the first portion

@@

			1,000,000 lines per batch. For each batch, we performed either one or two epochs.

@@

			we cannot generate the explanations directly from the input itself. However, because

@@

			the input of the terminus model would be. We generate intermediate data by first

@@

			terminus, and for each sub-model, running a prediction on the given input features, so
			that we have the intermediate results of each sub-model before the processing of the
			terminus model. We stored the results of the predictions of the sub-models and used
			for further analysis. This was done not only for the sample explanation data but also
			for a random sampling of the training data, so that the LIME explainer could use this
			data to generate localized estimated of model behavior.

@@

			We then summarized The results of the LIME explainer together on a per-feature basis
			and normalized them to produce a decimal number between -1 and 1 for every feature. In
			this case, a negative value implies that the given feature pushed the result of the
			model in the negative direction (i.e. not fake), while a positive value indicates the

@@

			previously described method together in a meaningful way in order to make a conclusion
			about the general usefulness of the individual features and the influence and
			potential use of the perceptible features. However, due to a lack of existing
			standardized metrics for the aggregate, global-scale evaluation of local explanations,
			we propose two metrics, with which we can better assess the usefulness of our
			method.

@@

	testing the surrogate model as well as generating many explanations for the purpose of
	aggregate evaluation. In this section, we will first discuss the dataset that we used for our
	experiments, the results and performance of the surrogate model, followed by an evaluation of
	the aggregated explanations.

@@

		The dataset used for these experiments is the In-the-Wild dataset, one which focuses on

@@

		aforementioned focus on generalization, its relative recency compared to some others and

@@

		We provide summary of some experiments based on the method described above in table
		\ref{table:eval-results}. The table details the following information: The type of
		terminus model used, where a "simple" terminus model does not include hidden layers; the
		set of features used, where "standard" is the feature set as described above including
		only one pitch-fluctuation feature, with expanded pitch fluctuations being the same
		feature set with the addition of more pitch fluctuation features with different comparison
		distances; the training batch size and the number of epochs per batch if there were more
		than one; followed by several performance metrics. Figures \ref{fig:roc_cterm},
		\ref{fig:roc_mpf} and \ref{fig:roc_mpf_cterm} depict the ROC curve of three of the models,
		which were selected for use in explanation generation in the following section.

@@

			In order to contextualize our results, we compare them to other results produced using
			tests on the same dataset. Table \ref{table:other_results} presents some experimental
			results that other methods have achieved on this dataset, sorted by source and model
			architecture. The results are given in EER, and the best result is in bold. In
			comparison to the other methods, our method performs very well, only beaten out by one
			other \cite{ranjan_statnet_2022}, which uses an expanded RawNet2-based network, taking
			the raw waveform as an input. They trained their method on 70\% of the dataset, used
			10\% for validation and the remaining 20\% for testing. This indicates a potential for
			raw-waveform based learned feature in the accurate detection of audio deepfakes, but
			does not offer a clear path for explainability, as the model does not make any
			intrinsic distinction of audio characteristics that humans are able to perceive. Our
			model remains, however, competitive, indicating that it can provide a good base for
			trustworthy explanations.

@@

		on the first 500 samples of the testing data. The models selected were the model with the
		best accuracy and EER results, with the standard features and with three hidden layers in
		the terminus model, the model with expanded pitch-fluctuation features and a simple
		terminus, which achieved the best AUC result, as well as the model with expanded
		pitch-fluctuation features and a terminus with three hidden layers. We selected these
		models were selected because we wanted to evaluate the best performing models, as well as
		see how the changed pitch-fluctuation feature affected the evaluation of the explanations.
		We then aggregated and summarized the explanations generated using both of the previously

@@

		presented in tables \ref{table:exp-results-cterm},
		\ref{table:exp-results-more-pitch-flucs} and \ref{table:exp-results-mpf-cterm}. The same
		statistics are also represented graphically in the figures \ref{fig:exp_cterm},
		\ref{fig:exp_mpf} and \ref{fig:exp_mpf_cterm} respectively.

@@

		features had either little, no or negative influence on the classification. Jitter
		features are, however, still notable for being perceptible features with a positive trust
		value.

@@

		more harm than good in the tested classifications. In this example, both jitter and
		shimmer features are notable for their positive trust values indicating their potential as
		trustworthy perceptible features for explanations.

@@

		examples, the trust value of the intensities feature was negative here. The trust of the
		jitter and shimmer features, however, remain positive.

@@

		for results more presentable to an end user. As can be seen in the previous section in

@@

		net-negative impact on the correctness of the classifications in the experiments,
		according to our trust metric, leaving the question open, if future experiments may
		perform better without these features. In the case of the model without expanded
		pitch-fluctuations, four perceptible features, f0 lengths, onset strengths, intensities
		and pitch fluctuations had an overall average influence on the classification of 0,
		measured on both metrics, indicating that the model did not learn a meaningful correlation
		between these features and the authenticity of audio samples in the dataset. These
		features did have an impact on the performance of the model with expanded pitch
		fluctuations, but the difference was still minimal. In the case of perceptible features
		with a positive trust score, the potential remains for their potential use in
		understandable explanations as, even though the classification needle of the model was not
		moved significantly by these features, they still demonstrate a certain reliability. In
		spite of this, we believe that it is still worth investigating if other kinds of
		perceptible features will have more of an influence on the overall classification or if
		there are other imperceptible features that could be used in the place of MFCCs that may
		provide a better balance between classification correctness and a proportional influence
		on the end result, so that the usefulness of the explanations can be even stronger than
		the method using the above feature combinations, and we believe that this kind of model
		architecture paired with LIME and evaluated with the presented metrics has the potential
		to be a backbone for future research.

@@

		usability of such explanations open, in addition to the question of what additional
		perceptible features can possibly be extracted to mirror what people have specifically
		identified in fake voice samples. Finally, we have not addressed in our experiments how

@@

	consistent and correct classifications. The explanations, however, placed most of the weight
	on the selected imperceptible features, with most perceptible features either having little,
	no or negative impact on the end classification with the exception of jitter and shimmer,
	which offer potential to be used in explanations.
